---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# riskintroanalysis

<!-- badges: start -->
<!-- badges: end -->

The riskintroanalysis R package provides functions to analyse the risk of introduction of animal diseases within a geographic area. It is intended to be useful as a stand-alone package, but also to integrate directly into the riskintro Rshiny app. The motivation behind these projects is to easily conduct geospatial risk analysis using existing data from WAHIS (World Animal Health Information System).

## Installation

You can install the development version of riskintroanalysis like so:

``` r
remotes::install_github("riskintroanalysis")
```


## Dev notes!!

Functions required:
  
  1. Calculate Risk of Introduction score from emission risk factors.
  2. Functions for validation dataset
  3. Functions for easily tidying datasets so they are valid
  4. Functions to go through each analysis (listed below) using validated datasets

See analysis examples to develop.

## Analysis

The package provides functions for certain risk of introduction analysis, but requires external data to be provided. The functions are inteded to be flexible, but each analytical method requires the corresponding data.

The RShiny app [riskintro](https://github.com/ardata-fr/riskintro) provides a graphical interface for the following analyses.

The central datasets to each method are: 

  1. **The emission risk score table**, which provides a risk of emission diseases (that is to say the risk of a disease spreading from that country). This score is based on summary of risk factors, many of which are available from WAHIS (see more [here](insert link to riksintrodata package docs)). This data exists for many countries around the world, and is augmented with geospatial data corresponding to each country's administrative boundaries.
  2. **The area of interest**, Geospatial data defining the borders of a region. This is the area for which we want to know the risk of introduction of a disease. The most common data is that of a country and its administrative boundaries, including internal boundaries. For example, French departments, or German states.
  
These two datasets are compared against each other through the following analyses:

  1. **Border lengths**: using the length of shared borders the risk of introduction is weighed comared to each neighbouring country and their risk score.
  2. **Border entry points**: using an additional dataset that defines broder entry points, risk of introduction is weighted by legality of entry point.
  3. **Animal mobility**: using an additional dataset defining legal animal commerce flows, risk of introduction is weighed based on the number of animals entering the area.
  4. **Road access risk**: using raster data of the world, applied to the area of interest, risk of introdcution is infered through road acces. An area more accessible by road is considered higher risk.
  
  These methods of analysis are intended to be used together or individually to montor is risk of introduction.


# Example 1: Tunisia

## Core datasets

### Emission Risk Factors:

Emission risk factors are used to create an overall emission risk value that is 
associated with the possible spread of a disease from that country. It is made up
of multiple scores which are weighted.

Here, we create two entries in the emission risk table for Algeria and Libya. The 
absence of government oversight in each area represents an increased risk of emission.
Therefore a 1 means there is additional risk associated with this score, and a 0
means no additional risk.

```{r}

library(dplyr)
library(riskintroanalysis)
library(sf)

algeria <- erf_row(
  iso3 = "DZA",
  country = "Algeria",
  disease = "Avian infectious laryngotracheitis",
  animal_category = "Domestic",
  species = "Birds",
  disease_notification = 0,
  targeted_surveillance = 1,
  general_surveillance = 0,
  screening = 1,
  precautions_at_the_borders = 1,
  slaughter = 1,
  selective_killing_and_disposal = 1,
  zoning = 1,
  official_vaccination = 1,
  last_outbreak_end_date = as.Date("30/06/2023"),
  commerce_illegal = 0L,
  commerce_legal = 0L
)

libya <- erf_row(
  iso3 = "LBY",
  country = "Libya",
  disease = "Avian infectious laryngotracheitis",
  animal_category = "Domestic",
  species = "Birds",
  disease_notification = TRUE,
  targeted_surveillance = 1,
  general_surveillance = 0,
  screening = 1,
  precautions_at_the_borders = 0,
  slaughter = 1,
  selective_killing_and_disposal = 1,
  zoning = 1,
  official_vaccination = 1,
  last_outbreak_end_date = as.Date("30/06/2019"),
  commerce_illegal = 0L,
  commerce_legal = 1
)

wahis_erf <- get_wahis_erf(
    disease = "Avian infectious laryngotracheitis",
    animal_category = "Domestic",
    species = "Birds"
  )

emission_risk_factors <- dplyr::bind_rows(
  algeria, 
  libya,
  wahis_erf
  )

emission_risk_table <- calc_weighted_emission_risk(emission_risk_factors = emission_risk_factors)

hist(
  emission_risk_table$emission_risk,
  main = "Histogram of emission risk score",
  xlab = "Emission risk scores"
)
```


It is also useful to join the world_sf geospatial dataset to emission risk for later.

```{r}

world <- riskintrodata::world_sf

emission_risk_sf <- dplyr::left_join(
  x = world, y = emission_risk_table, 
  by = c("iso3" = "iso3")
)

plot(select(emission_risk_sf, emission_risk), 
     main = "Emission risk scores")

```


### Area of interest (epidemiological units)

Epidemiological units (or epi units for short) is the term used for geographical areas of interests for analysing understand the risk of introduction. This data has no defaults and needs to be 
provided by the user, however packages like `geodata` can be used to get them. For this example, a shapefile containing the administrative boundaries of Tunisia exists in the `riskintrodata` package for demonstration purposes.

The Tunisia epidemiological units dataset will be essential to the following analysis methods.

```{r}
library(sf)

# Example with raw sf files, previously downloaded with geodata::gadm()
tunisia_raw <- read_sf(system.file(
  package = "riskintrodata",
  "samples", "tunisia", "epi_units", "tunisia_adm2_raw.gpkg"
   ))

# Apply mapping to prepare colnames and validate dataset
tunisia <- apply_mapping(
  tunisia_raw,
  mapping = mapping_epi_units(
    eu_id = NULL,
    eu_name = "NAME_2",
    user_id = NULL,
    geometry = "geom"
  ),
  validate = TRUE
)

plot(sf::st_geometry(tunisia))
```



## Analysis

The 4 methods of analysis provided by the package are detailed below:

This an example of analysing the risk of introduction for Tunisian governorates.

### Risk of introduction through land borders:

Required datasets:

1.  epi units
2.  internal neighbours table
3.  emission risk table

The border length method calculates the risk of introduction of an animal disease 
for a certain species based on shared land borders. In this example, we see Tunisia
shares land borders with Algeria and Libya (this is why entries of these two countries
were added to the emission risk table above). 

Risk of introduction through land borders is the risk of emission from of bordering countries weighted by the length of the shared border. If an epi unit only has one bordering country with emission risk X, then the epi units risk of introduction is also X. If two bordering countries exist with emission risk Y ans Z, then the risk of introduction is a weighted average of Y and Z, with more importance given to the 
country with the longer border.

The analysis method starts with finding shared borders between epi units and bordering countries, which requires an algoithm align borders.

```{r}
library(riskintroanalysis)
library(sf)
library(dplyr)

emission_risk_table

tun_files <-
  list.files(
    system.file(
      package = "riskintrodata",
      "samples",
      "tunisia",
      "animal_mobility"
    ),
    full.names = TRUE
  )

tun <- sf::read_sf("~/TUNISIA-EXAMPLE-01/UniteEPI_ADM/TUN_ADM2.shp") |> 
  select(
    eu_id = fid,
    eu_name = shapeName,
    geometry = geometry
  )

tun_neighbours <- riskintro:::neighbours_table |> 
  filter(country_id == "TUN")

bordering_countries <- world |> 
  filter(iso3 %in% tun_neighbours$neighbour_id)

# Run function to get shared borders
shared_borders <- calc_shared_border_lengths(
  epi_units = tun,
  eu_id_col = "eu_id",
  bordering_countries = bordering_countries,
  bc_id_col = "iso3"
)

ggplot() +
  geom_sf(data = bordering_countries, colour = "red", fill = "pink", alpha = 0.5) +
  geom_sf(data = tun, fill = "lightblue", colour = "blue", alpha = 0.5) +
  geom_sf(data = shared_borders, colour = "black") +
  geom_sf_label(data = shared_borders, aes(label = round(border_length)), size = 1.5) +
  coord_sf(xlim = c(7, 12), ylim = c(30,38), expand = FALSE) + 
  labs(
    title = "Output from get_shared_border_lengths() function "
  )

```
The main purpose of the function is to correct misaligned borders, as can be seen in
the overlaps and divergent borders. The epi units data is used as the "true" border, as
geospatial data from different sources are rarely well aligned.

A border length has been calculated for each epi unit that a foreign country. 
For most of these areas, the risk of introduction will simply equal the emission risk
of the neighbouring country, with the exception of Tunisia's southern most governorate. Which will have an emission risk weighed by the border lengths or each neighbour.

```{r}
eu_border_risk_intro <- calc_weighted_border_risk(
  epi_units = tun, 
  shared_borders = shared_borders,
  emission_risk = emission_risk_table
  )

plot(select(eu_border_risk_intro, border_risk) |> 
       mutate(border_risk = as.factor(round(border_risk, 1))), 
     main = "Weighted risk of introduction useing border method")

```



### Entry point method:

```{r}
library(riskintroanalysis)
library(riskintrodata)

entry_points_fp <-
  list.files(
    system.file(
      package = "riskintrodata",
      "samples",
      "tunisia",
      "entry_points"
    ),
    full.names = TRUE
  )

entry_points <- readr::read_csv(entry_points_fp)

entry_points <- apply_mapping(
  dataset = entry_points, 
  mapping = mapping_entry_points(
    point_name = "NAME", 
    lng = "LONGITUDE_X", 
    lat = "LATITUDE_Y", 
    mode = "MODE", 
    type = "TYPE", 
    sources = "SOURCES"
  ), 
  validate = TRUE
)

calc_entry_point_intro_risk(
  entry_points = entry_points |> select(-sources) |> distinct(), 
  entry_point_sources =  entry_points |> select(entry_points)
)



```

### Animal mobility method:

Considerations:
  1. fix exports, need to export as much as possible.
  1. add "countries" package as dep and re-export country_name
  1. Provide tools to tidy data for analysis and validate datasets 
  at the start of each function.
  1. Do we continue to include lealfet labels in output?
  1. create plot.generics for each output to a basic plot or leaflet?
  1 Combine calc_animal_mobility_point_risk and calc_animal_mobility_eu_risk into one function?

```{r}
library(tidyverse)
library(sf)
library(riskintro)
library(countries)

url <- "~/TUNISIA-EXAMPLE-01/Mobilite_animale/ANIMAL_MOBILITY.csv"
erf <- riskintro:::emission_risk_factor_defaults
tun <- sf::read_sf("~/TUNISIA-EXAMPLE-01/UniteEPI_ADM/TUN_ADM2.shp") |> 
  select(
    eu_id = fid,
    eu_name = shapeName,
    geometry = geometry
  )

erf <- filter(erf, disease == "Anthrax")
erf <- get_weighted_emission_risk(erf, riskintro:::emission_risk_weights)

animal_mobility <- read_csv2(
  url,
  col_types = cols(
    ORIGIN_NAME = col_character(),
    ORIGIN_COUNTRY = col_character(),
    ORIGIN_LONGITUDE_X = col_character(),
    ORIGIN_LATITUDE_Y = col_number(),
    DESTINATION_NAME = col_character(),
    DESTINATION_COUNTRY = col_character(),
    DESTINATION_LONGITUDE_X = col_character(),
    DESTINATION_LATITUDE_Y = col_number(),
    HEADCOUNT = col_double()
  ))

colnames(animal_mobility) <- c(
  "o_name",
  "o_country",
  "o_lng",
  "o_lat",
  "d_name",
  "d_country",
  "d_lng",
  "d_lat",
  "quantity"
)

animal_mobility <- animal_mobility |> 
  mutate(o_iso3 = countries::country_name(x = o_country, to = "ISO3"),
         d_iso3 = countries::country_name(x = d_country, to = "ISO3"),
         .before = 1)

analysis_point <- calc_animal_mobility_point_risk(
  animal_mobility = animal_mobility, 
  emission_risk = erf, 
  country_iso3 = "TUN"
    )

analysis_output <- calc_animal_mobility_eu_risk(
  animal_mobility_points = analysis_point, 
  epi_units = tun, 
  method = "mean"
  )

ggplot() +
  geom_sf(data = analysis_output, aes(geometry = geometry)) +
  geom_sf(data = analysis_point)

```

### Road accessibility method:

```{r}
# library(riskintroanalysis)


```
