---
output:
  github_document:
    toc: true
    toc_depth: 4
    number_sections: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# riskintroanalysis

The riskintroanalysis R package provides functions to analyse the risk of introduction of animal diseases within a geographic area. It is intended to be useful as a stand-alone package, but also to integrate directly into the riskintro Rshiny app. The motivation behind these projects is to easily conduct geospatial risk analysis using existing data from WAHIS (World Animal Health Information System).

# Installation

You can install the development version of riskintroanalysis like so:

``` r
remotes::install_github("riskintroanalysis")
```

# Analysis

The package provides functions for certain risk of introduction analysis, but requires external data to be provided. The functions are inteded to be flexible, but each analytical method requires the corresponding data.

The RShiny app [riskintro](https://github.com/ardata-fr/riskintro) provides a graphical interface for the following analyses.

The central datasets to each method are: 

  1. **The emission risk score table**, which provides a risk of emission diseases (that is to say the risk of a disease spreading from that country). This score is based on summary of risk factors, many of which are available from WAHIS (see more [here](insert link to riksintrodata package docs)). This data exists for many countries around the world, and is augmented with geospatial data corresponding to each country's administrative boundaries.
  2. **The area of interest**, Geospatial data defining the borders of a region. This is the area for which we want to know the risk of introduction of a disease. The most common data is that of a country and its administrative boundaries, including internal boundaries. For example, French departments, or German states.
  
These two datasets are used along with other data to calculate risk scores for the following analyses:

  1. **Border lengths**: using the length of shared borders the risk of introduction is weighed comared to each neighbouring country and their risk score.
  2. **Border entry points**: using an additional dataset that defines broder entry points, risk of introduction is weighted by legality of entry point.
  3. **Animal mobility**: using an additional dataset defining legal animal commerce flows, risk of introduction is weighed based on the number of animals entering the area.
  
The forth analysis method does not use emission risk data, just data from a road accesibility raster and the epidemiological units:  

  4. **Road access risk**: using raster data of the world, applied to the area of interest, risk of introdcution is infered through road acces. An area more accessible by road is considered higher risk.
  
These methods of analysis are intended to be used together or individually to montor is risk of introduction. This is done by ensuring risks are using the same scale (for example between 0 and 100) and they can easily added to a table of risk.
  
Additionally, some tools are provided to add other risk (such as the ones not show above). Currently supported are pre-calculated risks or risks from raster files. Which can be rescaled and added to the risk table as well. This allows of the risk analyst the flexibility to add other risks outside of the default risks.

# Example 1: Tunisia

`riskintroanalysis` is intended to be used alongside `sf` geospatial data manipulation, 
`dplyr` for general data manipulation, and `terra` for raster data.

```{r include=FALSE}
library(sf)
library(dplyr)
```

```{r echo=FALSE}
library(sf)
library(dplyr)
```


## Core datasets

The two core datasets are emission risk factors and epidemiological units. The
first is used in three out of four analysis methods, the second is used in all of them.

### Emission Risk Factors:

Emission risk factors are used to create an overall emission risk value that is 
associated with the possible spread of a disease from that country. It is made up
of multiple scores which are weighted.

Here, we create two entries in the emission risk table for Algeria and Libya. The 
absence of government oversight in each area represents an increased risk of emission.
Therefore a 1 means there is additional risk associated with this score, and a 0
means no additional risk.

```{r emission-risk-loading}

library(dplyr)
library(riskintrodata)
library(riskintroanalysis)
library(sf)

algeria <- erf_row(
  iso3 = "DZA",
  country = "Algeria",
  disease = "Avian infectious laryngotracheitis",
  animal_category = "Domestic",
  species = "Birds",
  disease_notification = 0,
  targeted_surveillance = 1,
  general_surveillance = 0,
  screening = 1,
  precautions_at_the_borders = 1,
  slaughter = 1,
  selective_killing_and_disposal = 1,
  zoning = 1,
  official_vaccination = 1,
  last_outbreak_end_date = as.Date("30/06/2023"),
  commerce_illegal = 0L,
  commerce_legal = 0L
)

libya <- erf_row(
  iso3 = "LBY",
  country = "Libya",
  disease = "Avian infectious laryngotracheitis",
  animal_category = "Domestic",
  species = "Birds",
  disease_notification = TRUE,
  targeted_surveillance = 1,
  general_surveillance = 0,
  screening = 1,
  precautions_at_the_borders = 0,
  slaughter = 1,
  selective_killing_and_disposal = 1,
  zoning = 1,
  official_vaccination = 1,
  last_outbreak_end_date = as.Date("30/06/2019"),
  commerce_illegal = 0L,
  commerce_legal = 1
)

wahis_erf <- get_wahis_erf(
    disease = "Avian infectious laryngotracheitis",
    animal_category = "Domestic",
    species = "Birds"
  )

emission_risk_factors <- dplyr::bind_rows(
  algeria, 
  libya,
  wahis_erf
  )

emission_risk_table <- calc_emission_risk(emission_risk_factors = emission_risk_factors)

hist(
  emission_risk_table$emission_risk,
  main = "Histogram of emission risk score",
  xlab = "Emission risk scores"
)
```


It is also useful to join the world_sf geospatial dataset to emission risk for later.

```{r display-world}
world <- riskintrodata::world_sf

emission_risk_sf <- dplyr::left_join(
  x = world, y = emission_risk_table, 
  by = c("iso3" = "iso3")
)

plot(select(emission_risk_sf, emission_risk), 
     main = "Emission risk scores")

```


### Epidemiological Units

Epidemiological units (or epi units for short) is the term used for geographical areas of interests for analysing understand the risk of introduction. This data has no defaults and needs to be 
provided by the user, however packages like `geodata` can be used to get them. For this example, a shapefile containing the administrative boundaries of Tunisia exists in the `riskintrodata` package for demonstration purposes.

The Tunisia epidemiological units dataset will be essential to the following analysis methods.

```{r epi-units}
library(sf)
# Example with raw sf files, previously downloaded with geodata::gadm()
tunisia_raw <- read_sf(system.file(
  package = "riskintrodata",
  "samples", "tunisia", "epi_units", "tunisia_adm2_raw.gpkg"
   ))

# Apply mapping to prepare colnames and validate dataset
tunisia <- apply_mapping(
  tunisia_raw,
  mapping = mapping_epi_units(
    eu_name = "NAME_2",
    geometry = "geom"
  ),
  validate = TRUE
)

plot(sf::st_geometry(tunisia))
```
What's left after using `apply_mapping()` is the required "epi_units" dataset that
is compatible with the rest of the analysis going forward.

## Analysis

The 4 methods of analysis provided by the package are detailed below:

This an example of analysing the risk of introduction for Tunisian governorates.
Based off of the core datasets

### Border risk

Required datasets:

1.  epi units
2.  internal neighbours table
3.  emission risk table

The border length method calculates the risk of introduction of an animal disease 
for a certain species based on shared land borders. In this example, we see Tunisia
shares land borders with Algeria and Libya (this is why entries of these two countries
were added to the emission risk table above). 

Risk of introduction through land borders is the risk of emission from of bordering countries weighted by the length of the shared border. If an epi unit only has one bordering country with emission risk X, then the epi units risk of introduction is also X. If two bordering countries exist with emission risk Y ans Z, then the risk of introduction is a weighted average of Y and Z, with more importance given to the 
country with the longer border.

The analysis method starts with finding shared borders between epidemiological units and bordering countries, which requires an algorithm align borders.

```{r shared-borders, cache=TRUE}
# Run function to get shared borders
shared_borders <- calc_border_lengths(
  epi_units = tunisia
)

# Get bordering countries to display on ggplot
# (This is internal to calc_border_lengths() function)
bordering_countries <- riskintrodata::neighbours_table |> filter(country_id == "TUN") |> 
  right_join(riskintrodata::world_sf, by = c("neighbour_id" = "iso3"))
            
ggplot() +
  geom_sf(data = bordering_countries, colour = "red", fill = "pink", alpha = 0.5, aes(geometry = geometry)) +
  geom_sf(data = tunisia, fill = "lightblue", colour = "blue", alpha = 0.5) +
  geom_sf(data = shared_borders, colour = "black") +
  geom_sf_label(data = shared_borders, aes(label = sprintf("%.fkm", border_length)), size = 1.5) +
  coord_sf(xlim = c(7, 12), ylim = c(30,38), expand = FALSE)
```

The main purpose of the function is to correct misaligned borders, as can be seen in the overlaps and divergent borders. The epidemiological units data is used for the "true" border as
geospatial data from different sources are rarely well aligned.

A border length has been calculated for each epi unit that a foreign country. For most of these areas, the risk of introduction will simply equal the emission risk of the neighbouring country, with the exception of Tunisia's southern most governorate. Which will have an emission risk weighed by the border lengths or each neighbour.

```{r border-risk-analysis}
ri_borders <- calc_border_risk(
  epi_units = tunisia, 
  shared_borders = shared_borders,
  emission_risk = emission_risk_table
  )

plot_risk(ri_borders)
```

We can see that Remada has a risk of introduction weighted by border lengths. The 
numbers shown along the borders are the emission risk weighted by border lengths 
with neighbouring countries of Algeria and Libya. As on Remada shares borders with both countries, it is the only one with a weighted average.

### Entry point risk

```{r entry-points-analysis}
library(riskintroanalysis)
library(dplyr)

entry_points_fp <-
    system.file(
      package = "riskintrodata",
      "samples",
      "tunisia",
      "entry_points", "BORDER_CROSSING_POINTS.csv"
    )

entry_points <- readr::read_csv(entry_points_fp)

entry_points <- apply_mapping(
  dataset = entry_points, 
  mapping = mapping_entry_points(
    point_name = "NAME", 
    lng = "LONGITUDE_X", 
    lat = "LATITUDE_Y",
    mode = "MODE", 
    type = "TYPE", 
    sources = "SOURCES"
  ), 
  validate = TRUE
)

ri_entry_points <- calc_entry_point_risk(
  entry_points = entry_points,
  epi_units = tunisia,
  emission_risk = emission_risk_table
)

plot_risk(ri_entry_points)
```

All epi units containing an entry point (with an associated weighted emission risk) now have a risk of introduction based on the points within its area.

### Animal mobility

```{r animal-mobility-analysis}
library(riskintroanalysis)
animal_mobility_fp <- system.file(package = "riskintrodata", "samples", "tunisia", "animal_mobility", "ANIMAL_MOBILITY_raw.csv")

animal_mobility_raw <- readr::read_csv(animal_mobility_fp)

animal_mobility <- apply_mapping(
  animal_mobility_raw,
  mapping = mapping_animal_mobility(
    o_name = "ORIGIN_NAME",
    o_lng = "ORIGIN_LONGITUDE_X",
    o_lat = "ORIGIN_LATITUDE_Y",
    d_name = "DESTINATION_NAME",
    d_lng = "DESTINATION_LONGITUDE_X",
    d_lat = "DESTINATION_LATITUDE_Y",
    quantity = "HEADCOUNT"
  )
)

ri_animal_mobility <- calc_animal_mobility_risk(
  animal_mobility = animal_mobility, 
  emission_risk = emission_risk_table, 
  epi_units = tunisia,
  method = "mean"
)

plot_risk(ri_animal_mobility)
```

Some points remain grey indicating that the country from which the flow comes does
not have an entry in the emission risk factors table.

### Road accessibility

Road access risk is calculated from the road accessibility raster file, a global raster that contains data of distance from roads. This data is aggregated over each epidemiological unit of Tunisia.

```{r display-raster, asis = TRUE}
library(riskintroanalysis)
library(dplyr)
library(terra)

riskintrodata::init_riskintrodata_cache()
road_raster_fp <- riskintrodata::download_road_access_raster()
road_raster <- terra::rast(road_raster_fp)

ri_road_access <- calc_road_access_risk(
  epi_units = tunisia, 
  road_access_raster = road_raster, 
  aggregate_fun = "mean"
)

plot(
  extract_raster(ri_road_access), 
  main = "Raster cropped to epidemiological units"
)


plot_risk(ri_road_access)
```

This shows us what the raster data looks like for Tunisia. Now we will aggragate these values over each administrative area. Here, mean is used.

The output shows the road access risk associated with each epidemiological unit.

## Risk rescaling

Once all the analysis methods have given a result, we need to be able to compare them. Notably, road access risk is not on the same 0 to 12 scale as the other three analyses. You might also want to add your own risk scores using other methods.

To make sure we can compare across risks, riskintro proposes the `rescale_risk` function, whose primary purpose is to normalise risk scores to a range between 0 and 100. Its secondary role is to transform the risk distribution using quadratic, exponential or sigmoid functions (and inverse versions of these). Other options include reversing the scaling as well. 

A linear transformation looks like the following. This uses the maximum value of `road_access_risk` as the top of the range, whereas it could be better to take the global maximum in the raster if you are intending to comparing across countries later.

```{r rescaling-road-access-risk-1}
rescaled <- rescale_risk_scores(
  dataset = ri_road_access,
  cols = "road_access_risk",
  from = c(0, max(ri_road_access$road_access_risk)),
  to = c(0, 100),
  method = "linear", 
  names_to = "scaled_road_access_risk",
  keep_cols = TRUE
)
plot(rescaled$road_access_risk, rescaled$scaled_road_access_risk)
```

While a sigmoid transformation looks like the below. The choice of the `method` argument are left up to the judgement of the analyst.

Note that a risk dataset ouput from one of the 4 default risk analyses will detect
which column to scale. The `cols` and `names_to` arguments are not required.
```{r rescaling-road-access-risk-2}
ri_road_access_scaled <- rescale_risk_scores(
  data = ri_road_access,
  from = c(0, max(ri_road_access$road_access_risk)),
  method = "sigmoid", 
  keep_cols = FALSE
)
```


## Non-default risks 

Riskintro also provides more generic workflows for adding any sort of risk scores
to a risk table alongside the default risk.

### Risk from raster file

First, import or generate the raster data to use for analysis. This example generates fictitious water courses in Tunisia which could be a source of introduction risk.

```{r, warning=FALSE}
library(terra)
tun_ext <- ext(6, 12, 30, 38)
r_tun <- rast(tun_ext, res = 0.1, crs = "EPSG:4326")
wkt_lines <- c(
  "LINESTRING (9.0 36.0, 9.5 35.0, 10.0 34.5)",
  "LINESTRING (8.5 33.5, 9.0 33.0, 10.0 32.5)"
)
water_sv <- vect(data.frame(wkt = wkt_lines),
                 geom = "wkt",
                 crs  = "EPSG:4326")
water_sv$water <- 1
water_rast <- rasterize(
  water_sv,
  r_tun,
  field      = "water",
  background = 0
  )
plot(water_rast, main = "Example watercourses")
plot(tunisia, add = TRUE, color = "red")
```

```{r}
ri_water <- augment_epi_units_with_raster(
  epi_units = tunisia, 
  raster = water_rast, 
  risk_name = "water_risk", 
  aggregate_fun = "mean"
) |> 
  rescale_risk_scores(
    cols = "water_risk", 
    from = c(0,1),
    to = c(0,100), 
    method = "quadratic"
  )

ggplot(ri_water) +
  geom_sf(aes(fill = water_risk)) +
  scale_fill_viridis_c(limits = c(0, 100))
```
### Pre-calculated risk

In the case an analyst already has a dataset will risk scores for each epidemiolocal unit, it is possible to simply join this dataset to the risk table assuming that:

1.  a correct join key exists in the pre-calculated dataset and,
2.  there is no more an one risk score associated with each epidemiological unit.

Here we generate a dataset from a random uniform distribution

```{r}
ri_random <- tunisia |> 
  select(my_key = eu_id) |> 
  mutate(
    ri_random = runif(nrow(tunisia), min = 0, max = 100)
         ) |> 
  filter(row_number() <= 30) |> 
  rescale_risk_scores(
    cols = "ri_random",
    from = c(0, 100), to = c(0, 100), method = "sigmoid"
  )

```


## Summary table 

Once all the risk methods have been compiled and scaled similarly, it can be handy to have them all in one table. Each row is an epidemiological unit and its various risks scores.

```{r summary-table}
# initialise the risk_table 
rt <- risk_table(tunisia, scale = c(0, 100))
# Add risks that have already been rescaled
rt <- add_risk(risk_table = rt, risk_data = ri_road_access_scaled)
rt <- add_risk(risk_table = rt, risk_data = ri_water)
rt <- add_risk(risk_table = rt, risk_data = ri_random, join_by = "my_key")

# add other risks, rescaling them as needed
rt <- add_risk(
  risk_table = rt,
  risk_data = rescale_risk_scores(
    dataset = ri_entry_points, 
    method = "sigmoid"
    )
)

rt <- add_risk(
  risk_table = rt,
  risk_data = rescale_risk_scores(
    dataset = ri_borders
    )
)

rt <- add_risk(
  risk_table = rt,
  risk_data = rescale_risk_scores(
    dataset = ri_animal_mobility
    )
)
summarised_risks <- summarise_risk_scores(rt, method = "max")

plot_risk(summarised_risks)
```




