---
title: "Introduction risk"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction risk}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(riskintroanalysis)
library(riskintrodata)
library(dplyr)
library(sf)
library(ggplot2)
```

## Introduction

**Introduction risk** is the overall assessment of disease introduction probability for each epidemiological unit, combining multiple risk pathways into a single comprehensive score. This vignette demonstrates how to:

1. Calculate risks from multiple sources (border, entry points, animal mobility, road access)
2. Rescale risks to a common 0-100 scale
3. Combine risks into a unified risk table
4. Summarize risks using different aggregation methods
5. Visualize the final introduction risk

The key insight is that disease introduction can occur through multiple pathways, and a comprehensive risk assessment must account for all relevant routes.

## Preparing Core Data

All risk analyses require two core datasets: epidemiological units and (for most methods) emission risk scores.

### Epidemiological Units

```{r epi-units}
# Load and validate epidemiological units
tunisia_raw <- read_sf(system.file(
  package = "riskintrodata",
  "samples", "tunisia", "epi_units", "tunisia_adm2_raw.gpkg"
))

tunisia <- validate_dataset(
  x = tunisia_raw,
  table_name = "epi_units",
  eu_name = "NAME_2",
  geometry = "geom"
) |> extract_dataset()
```

### Emission Risk Scores

```{r emission-risk}
# Create emission risk factors for Tunisia's neighbors
algeria <- erf_row(
  iso3 = "DZA",
  country = "Algeria",
  disease = "Avian infectious laryngotracheitis",
  animal_category = "Domestic",
  species = "Birds",
  disease_notification = 0,
  targeted_surveillance = 1,
  general_surveillance = 0,
  screening = 1,
  precautions_at_the_borders = 1,
  slaughter = 1,
  selective_killing_and_disposal = 1,
  zoning = 1,
  official_vaccination = 1,
  last_outbreak_end_date = as.Date("2023-06-30"),
  commerce_illegal = 0L,
  commerce_legal = 0L
)

libya <- erf_row(
  iso3 = "LBY",
  country = "Libya",
  disease = "Avian infectious laryngotracheitis",
  animal_category = "Domestic",
  species = "Birds",
  disease_notification = TRUE,
  targeted_surveillance = 1,
  general_surveillance = 0,
  screening = 1,
  precautions_at_the_borders = 0,
  slaughter = 1,
  selective_killing_and_disposal = 1,
  zoning = 1,
  official_vaccination = 1,
  last_outbreak_end_date = as.Date("2019-06-30"),
  commerce_illegal = 0L,
  commerce_legal = 1
)

wahis_erf <- get_wahis_erf(
  disease = "Avian infectious laryngotracheitis",
  animal_category = "Domestic",
  species = "Birds"
)

emission_risk_factors <- dplyr::bind_rows(algeria, libya, wahis_erf)
emission_risk_table <- calc_emission_risk(emission_risk_factors = emission_risk_factors)
```

## Calculating Individual Risks

Each risk pathway is calculated separately using its specific method and data requirements. Below we show the input data types needed for each analysis.

### Border Risk

**Input data types:**
- Epidemiological units (spatial vector - sf polygon)
- Emission risk table (tabular data frame)
- Shared borders are calculated internally from the epidemiological units

```{r border-risk}
shared_borders <- calc_border_lengths(epi_units = tunisia)

ri_borders <- calc_border_risk(
  epi_units = tunisia,
  shared_borders = shared_borders,
  emission_risk = emission_risk_table
)
```

### Entry Point Risk

**Input data types:**
- Entry points (tabular CSV with coordinates - converted to spatial points internally)
- Epidemiological units (spatial vector - sf polygon)
- Emission risk table (tabular data frame)

```{r entry-point-risk}
entry_points_fp <- system.file(
  package = "riskintrodata",
  "samples", "tunisia", "entry_points", "BORDER_CROSSING_POINTS.csv"
)

entry_points <- readr::read_csv(entry_points_fp, show_col_types = FALSE)

entry_points <- validate_dataset(
  x = entry_points,
  table_name = "entry_points",
  point_name = "NAME",
  lng = "LONGITUDE_X",
  lat = "LATITUDE_Y",
  mode = "MODE",
  type = "TYPE",
  sources = "SOURCES"
) |> extract_dataset()

ri_entry_points <- calc_entry_point_risk(
  entry_points = entry_points,
  epi_units = tunisia,
  emission_risk = emission_risk_table
)
```

### Animal Mobility Risk

**Input data types:**
- Animal mobility flows (tabular CSV with origin and destination coordinates)
- Epidemiological units (spatial vector - sf polygon)
- Emission risk table (tabular data frame)

```{r animal-mobility-risk}
animal_mobility_fp <- system.file(
  package = "riskintrodata",
  "samples", "tunisia", "animal_mobility", "ANIMAL_MOBILITY_raw.csv"
)

animal_mobility_raw <- readr::read_csv(animal_mobility_fp, show_col_types = FALSE)

animal_mobility <- validate_dataset(
  x = animal_mobility_raw,
  table_name = "animal_mobility",
  o_name = "ORIGIN_NAME",
  o_lng = "ORIGIN_LONGITUDE_X",
  o_lat = "ORIGIN_LATITUDE_Y",
  d_name = "DESTINATION_NAME",
  d_lng = "DESTINATION_LONGITUDE_X",
  d_lat = "DESTINATION_LATITUDE_Y",
  quantity = "HEADCOUNT"
) |> extract_dataset()

ri_animal_mobility <- calc_animal_mobility_risk(
  animal_mobility = animal_mobility,
  emission_risk = emission_risk_table,
  epi_units = tunisia,
  method = "mean"
)
```

### Road Access Risk

**Input data types:**
- Road accessibility raster (raster file - global dataset)
- Epidemiological units (spatial vector - sf polygon)
- Note: This method does NOT require emission risk scores

```{r road-access-risk, eval=FALSE}
# Note: This step downloads a large raster file (~500 MB)
library(terra)

riskintrodata::init_riskintrodata_cache()
road_raster_fp <- riskintrodata::download_road_access_raster()
road_raster <- terra::rast(road_raster_fp)

ri_road_access <- calc_road_access_risk(
  epi_units = tunisia,
  road_access_raster = road_raster,
  aggregate_fun = "mean"
)
```

```{r fake-road-risk, echo=FALSE}
# For vignette building, create synthetic road access data
ri_road_access <- tunisia |>
  mutate(road_access_risk = runif(n(), 0, 50))
attr(ri_road_access, "risk_col") <- "road_access_risk"
attr(ri_road_access, "scale") <- c(0, 50)
attr(ri_road_access, "table_name") <- "road_access"
```

## Rescaling Risks to a Common Scale

Before combining risks, we must ensure they're all on the same scale. The standard approach is to rescale all risks to 0-100.

Most risks start on a 0-12 scale (from emission risk scores), but road access uses its own scale based on the raster values.

```{r rescale-risks}
# Border risk: from 0-12 to 0-100
ri_borders_scaled <- rescale_risk_scores(
  dataset = ri_borders,
  from = c(0, 12),
  to = c(0, 100),
  method = "linear"
)

# Entry point risk: from 0-12 to 0-100
ri_entry_points_scaled <- rescale_risk_scores(
  dataset = ri_entry_points,
  from = c(0, 12),
  to = c(0, 100),
  method = "sigmoid"
)

# Animal mobility risk: from 0-12 to 0-100
ri_animal_mobility_scaled <- rescale_risk_scores(
  dataset = ri_animal_mobility,
  from = c(0, 12),
  to = c(0, 100),
  method = "linear"
)

# Road access risk: from 0-50 to 0-100
ri_road_access_scaled <- rescale_risk_scores(
  dataset = ri_road_access,
  from = c(0, max(ri_road_access$road_access_risk)),
  to = c(0, 100),
  method = "linear"
)
```

### Rescaling Methods

The `method` parameter controls how values are transformed:

- **linear** - Straightforward proportional scaling
- **quadratic** - Emphasizes higher values (x²)
- **exponential** - Strongly emphasizes higher values (e^x)
- **sigmoid** - S-curve that compresses extremes and expands middle values

Choose the method based on how you want to weight different risk levels in your analysis.

## Creating a Risk Table

A **risk table** is a unified structure that holds all risk scores for each epidemiological unit. It ensures all risks are on the same scale and provides a consistent way to manage multiple risk pathways.

```{r create-risk-table}
# Initialize the risk table with the 0-100 scale
rt <- risk_table(tunisia, scale = c(0, 100))

# Add each risk to the table
rt <- add_risk(risk_table = rt, risk_data = ri_borders_scaled)
rt <- add_risk(risk_table = rt, risk_data = ri_entry_points_scaled)
rt <- add_risk(risk_table = rt, risk_data = ri_animal_mobility_scaled)
rt <- add_risk(risk_table = rt, risk_data = ri_road_access_scaled)

# View the structure
head(rt)
```

The risk table now contains all four risk scores for each epidemiological unit, all on the 0-100 scale.

## Summarizing Risks

With all risks in one table, we can summarize them into a single overall introduction risk score using different aggregation methods:

```{r summarize-risks}
# Maximum risk across all pathways
risk_max <- summarise_scores(rt, method = "max")

# Mean risk across all pathways
risk_mean <- summarise_scores(rt, method = "mean")

# Median risk across all pathways
risk_min <- summarise_scores(rt, method = "min")
```

### Choosing an Aggregation Method

Different methods reflect different risk philosophies:

- **max** - "Weakest link" approach: overall risk equals the highest individual risk. Conservative choice that emphasizes the most dangerous pathway.
- **mean** - Averages all pathways equally. Good when all pathways contribute independently to risk.
- **min** - Rarely used, but available if you need the minimum risk value.

For disease introduction, **max** is often preferred because a single high-risk pathway can facilitate disease entry regardless of other pathways.

## Visualizing Introduction Risk

### Static Maps

Compare the different summarization methods visually:

```{r compare-methods, fig.width=10, fig.height=8}
library(patchwork)

p1 <- plot_risk(risk_max) + labs(title = "Maximum Risk")
p2 <- plot_risk(risk_mean) + labs(title = "Mean Risk")
p3 <- plot_risk(risk_min) + labs(title = "Minimum Risk")

p1 + p2 + p3
```

### Interactive Map

For detailed exploration, use an interactive map:

```{r interactive-map}
library(leaflet)

plot_risk_interactive(risk_max)
```

Click on any governorate to see its overall introduction risk score and individual pathway contributions.

## Manual Risk Overrides

In some cases, expert knowledge may suggest modifying risk scores for specific units. The risk table structure supports this:

```{r manual-override, eval=FALSE}
# Override risk for specific units based on expert judgment
rt_modified <- rt |>
  mutate(
    border_risk = case_when(
      eu_name == "Tunis" ~ 50,  # Reduce due to enhanced border controls
      TRUE ~ border_risk
    )
  )

# Recalculate summary with overrides
risk_max_modified <- summarise_scores(rt_modified, method = "max")
```

## Summary

This vignette demonstrated the complete workflow for assessing overall disease introduction risk:

1. **Calculate individual risks** from multiple pathways
2. **Rescale to common scale** (typically 0-100)
3. **Combine into risk table** for unified management
4. **Summarize using appropriate method** (max, mean, median)
5. **Visualize and explore** results

This approach provides a comprehensive, transparent assessment of disease introduction risk that accounts for multiple pathways while remaining flexible enough to incorporate expert judgment and additional risk factors.

## Further Reading

For detailed information on each risk calculation method, see:

- [Border Risk Analysis](border-risk-analysis.html)
- [Entry Points Analysis](entry-points-analysis.html)
- [Animal Mobility Analysis](animal-mobility-analysis.html)
- [Road Access Analysis](road-access-analysis.html)
- [Additional Risks](additional-risks.html)
- [Emission Scores](emission-scores.html)
