---
title: "Risk scaling of entry points"
format: 
  html:
    toc: true
    html-math-method: mathjax
execute: 
  echo: false
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Risk scaling of entry points}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
#| include: false
library(riskintroanalysis)
library(dplyr)
library(ggplot2)
library(tidyr)
```

## Introduction


__Entry points__ are one of the default risk factors in `riskintroanalysis`.
They correspond places in the territory known to be risk sources of introduction due to the passage of animals.
They typically correspond to border posts, or simply routes and paths that cross the border of the target territory, which are known to be used for animal transportation or transhumance.

They can also be interior (i.e. not in the border) locations such as airports or aerodromes.
But they are distinguished from the desintations of __animal movement__ in that:

1. they represent the introduction of animals outside formal trade flows,

2. the volume of animal flows are not known, nor traceable,

3. the sanitary regulations and controls are weaker or inexistent.

We distinguish __controlled__ and __not-controlled__ entry points.
For example, a border post with a sanitary control of transhumant herds can be considered a _controlled_ entry point.
By contrast, a mountain pass known to be used for unauthorised introduction of animals would be considered _not-controlled_ and typically would entail a higher risk of pathogen introduction.


## Input data

For each epidemiological unit, the relevant quantities considered are:

1. the number of entry points

2. their status (controlled/not-controlled)

3. the emission scores of the source countries

In practice, you need to provide the entry points with their location and their status.
Furthermore, for each entry point, you need to identify the countries that are potential sources of animal flows.


## Step 1: Point Exposure

![Scematic example of a epidemiological unit with two controlled and one not-controlled entry points.](img/diagram_entry-points.svg){#fig-diagram}

For each entry point, we first compute its value of _exposure_ $e$, which wheights the number and emission scores $\nu_k$ of the source countries.

$$
e = \sum_k \nu_k / \nu_{\text{max}}
$$ {#eq-exposure}
where $\nu_{\text{max}} = 12$ is the maximum possible emission score.

In the example in @fig-diagram, the corresponding exposure values for the entry points, top to bottom, are:

- $e_1 = (12 + 3)/12 = 1.25$

- $e_2 = (3 + 9 + 3)/12 = 1.25$

- $e_3 = 3/12 = 0.25$

These values can be interpreted as a _effective number of __risky__ sources_. 
The first entry point has 2 sources, but the second has a lower emission score, so it counts as a quarter of a _risky_ source.
The second entry point has 3 sources, but together, they accumulate to an effective equivalent equal to the exposure of the first entry point.
The bottom entry point has only one source of animals with a low emission score, resulting in only a quarter of a _risky_ source.

Note that the controlled/not-control status of the entry points do not affect their exposure values.

In summary @eq-exposure reflects the principle of _compensation_: the number of sources can balance out low emission scores.
This contrasts with the principle of _maximisation_, where the exposure level is equal to the highest emission score, regardless of the number of sources. 
There are good arguments for both principles, depending on the situation.
We might decide to implement the [LogSumExp](https://en.wikipedia.org/wiki/LogSumExp) function in the future as a compromise.


## Step 2: Effective number of controlled/not-controlled points

In the next step, we aggregate the exposures of the controlled and not-controlled entry points separately.
The resulting numbers can be interpreted as the _effective number of entry points_, controlled and not-controlled.

Once again, we choose the _addition_ as the aggregation function, although we can eventually switch to _max_ or _LogSumExp_.

For the unit depicted in @fig-diagram, this results in adding the exposure values of $e_1$ and $e_2$ together for the controlled entry points, whereas the only not-controlled entry point is left alone.
In total, the epidemiological unit is considered to have:

```{r}
xl_ex <- 2.5
xi_ex <- 0.25
```
- `r xl_ex` effective controlled entry points

- `r xi_ex` effective not-controlled entry points


## Step 3: Scaling the risk of introduction

Finally, we need to specify the map from these effective numbers of entry points to a risk scale (i.e. a value from 0 to 100, quantifying the plausibility of disease introduction in a given unit).

For this, we define the following scaling function:

```{r}
case_dat <- expand.grid(
  xl = seq(0, 4, length = 51),
  xi = seq(0, 4, length = 51)
)
```

```{r}
case_annotations <- data.frame(
  xl = c(0, 4),
  xi = c(4, 0),
  label = c("🡑M", "🠂M/𝜆"),
  hj = c(0, -.2),
  vj = c(-1, 0)
)
```

```{r}
#| label: fig-bivariate-scaling
#| fig.cap: "Bivariate risk scaling function of controlled and not-controlled entry points, with a point marker at the effective numbers of entry points in the example."
case_dat |>
  transform(
    risk = riskintroanalysis:::scale_entry_points(xl, xi, 3, 1, 1, 100)
  ) |> 
  ggplot(aes(xl, xi)) +
  geom_raster(
    aes(fill = risk)
  ) +
  geom_point(
    x = xl_ex, y = xi_ex,
    colour = 'grey90',
    size = 2
  ) +
  geom_text(
    data = case_annotations,
    aes(label = label, hjust = hj, vjust = vj)
  ) +
  coord_fixed(clip = 'off') +
  scale_fill_viridis_c(
    option = "D",
    limits = c(0, 100)
  ) +
  labs(
    x = "Controlled points",
    y = "Not-controlled points",
    fill = "Risk"
  ) +
  theme_minimal()

```


The corresponding mathematical expression is:

$$
f(x_c, x_u) = M s(\beta x_u + s^{-1}(s(\alpha x_c) / \lambda))
$$ {#eq-scaling}
where $s(x) = (1 - \exp(-x)) / (1 + \exp(-x))$ is a semi-sigmoid curve on the non-negative reals (@fig-sigmoid).

```{r}
#| label: fig-sigmoid
#| fig.width: 3
#| fig.height: 2
#| fig.cap: "Semi-sigmoid function $s(x)$."
ggplot() +
xlim(0, 5) +
geom_function(fun = riskintroanalysis:::sigmoid) +
theme_minimal() +
labs(x = NULL, y = NULL) +
theme(
  panel.grid.minor = element_blank(),
  panel.grid.major.x = element_blank()
)
```


For the running example from the previous section, the corresponding location for the epidemiological unit is indicated in @fig-bivariate-scaling with a dot, where the function takes the value of 
`r riskintroanalysis:::scale_entry_points(xl_ex, xi_ex, 3, 1, 1, 100) |> round(2)`.
This would be the risk of introduction attributed to entry points, associated to this particular epidemiological unit.

While the general shape of this bivariate smooth function (i.e., a surface) is fixed, you can set a few parameters to adjust the risk values.

The risk is fixed at 0 in the bottom-left corner.
I.e., when there are no entry points at all, either controlled or not, the risk of introduction due to this factor is 0.

The risk increases progressively when the number of entry points increase in either the horizontal direction of _controlled_ points, or in the vertical direction of _not-controlled_ points.
Moving up, the risk increases with the number of _not-controlled_ entry points up to an asymptotic limit $M$ (in @fig-bivariate-scaling, $M = 100$).
By contrast, in the absence of _not-controlled_ entry points, the risk increases with the number of _controlled_ points up to an asymptotic limit of $M/\lambda$, where $\lambda$ represents how much riskier _not-controlled_ points are, with respect to controlled ones. In @fig-bivariate-scaling, $\lambda = 3$.

Furthermore, you can control _how fast_ these asymptotic limits are approached with the parameters $\alpha > 0$ and $\beta > 0$.
In @fig-bivariate-scaling, $\alpha = \beta = 1$, resulting in risk values very close to the asymptotic limits after 3 or 4 entry points of each type.
For instance, the risk value at $(0, 4)$ is `r riskintroanalysis:::scale_entry_points(0, 4, 3, 1, 1, 100) |> round(2)`, very close to the limit $M = 100$ for infinite _non-controlled_ entry points.
On the other hand, the risk value at $(3, 0)$ is `r riskintroanalysis:::scale_entry_points(3, 0, 3, 1, 1, 100) |> round(2)`, very close to the limit $M/\lambda = 100 / 3$ for infinite _controlled_ entry points without any _non-controlled_ point (@fig-marginal-scaling).

```{r}
#| label: fig-marginal-scaling
#| fig.cap: "Marginal risk scaling function of controlled and uncontrolled entry points, with default and alternative values of parameters."
#| fig.height: 4
case_dat |>
  transform(
    risk_1 = riskintroanalysis:::scale_entry_points(xl, xi, 3, 1, 1, 100),
    risk_2 = riskintroanalysis:::scale_entry_points(xl, xi, 3, 1/2, 2, 100)
  ) |> 
  filter(xl == 0 | xi == 0) |>
  mutate(
    x = xl + xi,
    marg = factor(ifelse(xi == 0, "Controlled", "Uncontrolled"))
  ) |>
  pivot_longer(
    cols = starts_with("risk"),
    names_to = "case",
    names_prefix = "risk_",
    names_transform = factor,
    values_to = "risk"
  ) |>
  inner_join(
    data.frame(
      marg = rep(factor(c("Controlled", "Uncontrolled")), each = 2),
      case = rep(factor(1:2), times = 2),
      ab = c("1", "1/2", "1", "2")
    ),
    by = join_by(marg, case)
  ) |>
  inner_join(
    data.frame(
      case = factor(1:2),
      alpha_value = c(1, .4)
    ),
    by = 'case'
  ) |>
  # filter(x == max(x))
  # filter(x == 2) |>
  # mutate(lab = paste(rep(c("𝛼 =", "𝛽 ="), each = 2), ab))
  ggplot(aes(x, risk)) +
  geom_line(
    aes(alpha = alpha_value, group = case:marg),
    show.legend = FALSE
  ) +
  ## Controlled / Uncontrolled labels
  geom_text(
    data = \(.) filter(., x == max(x), ab == "1"),
    aes(label = marg, hjust = 1, vjust = 0),
    nudge_y = c(1, 5)
  ) +
  ## Parameter values alpha / beta
  geom_text(
    data = \(.) {
      filter(., x == 2) |>
      mutate(lab = paste(rep(c("𝛼 =", "𝛽 ="), each = 2), ab))
    },
    aes(label = lab, hjust = 1, vjust = 0, alpha = alpha_value),
    nudge_y = 1,
    show.legend = FALSE
  ) +
  labs(
    x = "Number of points",
    y = "Risk"
  ) +
  scale_alpha_continuous(limits = c(0, 1)) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )

```

If you considered, for instance, that one _non-controlled_ entry point is as bad as it can get, and beyond that there is not much of a difference, you would then increase the value of $\alpha$ as in @fig-marginal-scaling.
By contrast, you could decide that additional _controlled_ entry points increase risk more progressively, and it takes more entry points to make up for any considerable risk of introduction.
In this case, you could then decrease the value of $\beta$.

Note that the values of $\alpha$ and $\beta$ are independent from each other, as long as are both positive values.


## Equivalent effective number of uncontrolled entry points

For convenience, we can reparameterise the @eq-scaling into the following univariate function:

$$
g(x) = M s(\alpha x)
$$ {#eq-univariate-scaling}
where
$$
x = x_u + \hat{x}_u
$$
and,
$$
\hat{x}_u = s^{-1}(s(\beta x_c) / \lambda) / \alpha
$$

The last expression can be interpreted as the number of equivalent _uncontrolled_ entry points, for a given number of _controlled_ entry points.
@fig-equivalent-uncontrolled displays the curve for the parameters used previously ($\alpha = \beta = 1$; $\lambda = 3$).

```{r}
scale_controlled <- function(x, alpha, beta, lambda) {
  riskintroanalysis:::inv_sigmoid(
    riskintroanalysis:::sigmoid(beta * x) / lambda)  / alpha
}
```

```{r}
#| label: fig-equivalent-uncontrolled
#| fig.cap: "Equivalent uncontrolled entry points."
#| fig.width: 4
#| fig.height: 3
data.frame(
  x = seq(0, 5, length = 51)
) |>
  transform(
    y = scale_controlled(x, alpha = 1, beta = 1, lambda = 3)
  ) |>
  ggplot() +
  aes(x, y) +
  geom_line() +
  labs(
    x = expression("Controlled points"~x[c]),
    y = expression("Equivalent uncontrolled points"~hat(x)[u])
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )
```


```{r}
hatxu_ex <- scale_controlled(xl_ex, alpha = 1, beta = 1, lambda = 3)
(risk_ex <- 100 * riskintroanalysis:::sigmoid(1 * (xi_ex + hatxu_ex)))
```
With this equivalence established, we can consider $x$ as the _equivalent number of effective uncontrolled entry points_ for a given unit.
In the simple example depicted in @fig-diagram we had obtained $x_u = `r xi_ex`$ effective uncontrolled entry points and $x_c = `r xl_ex`$ effective controlled entry points, that are equivalent to $\hat{x}_u = `r round(hatxu_ex, 2)`$ uncontrolled points, in terms of risk.

In summary this unit would have $x = `r round(xi_ex + hatxu_ex, 2)`$ equivalent effective uncontrolled entry points, that we can run through the function @eq-univariate-scaling to evaluate an associated risk of 
`r round(risk_ex)`.

The interface will display the univariate risk scaling curve @eq-univariate-scaling, in terms of the equivalent effective number of uncontrolled entry points $x$, allowing to choose between several alternative shapes.
Points corresponding to the observed values of $x$ for all the epidemiological units will be superposed for reference.
The location of these points will vary, as a function of the selected parameters $\alpha > 0$, $\beta > 0$ and $\lambda > 1$.

```{r}
#| label: fig-scaling-univariate
#| fig.cap: "User interface for the univariate scaling of entry points."
#| fig.width: 4
#| fig.height: 3
data.frame(
  x = seq(0, 3, length = 51)
) |>
  transform(
    y = 100 * riskintroanalysis:::sigmoid(1 * x)
  ) |>
  ggplot() +
  aes(x, y) +
  geom_line() +
  geom_point(
    x = xi_ex + hatxu_ex,
    y = risk_ex,
    size = 4,
    alpha = 0.5
  ) +
  labs(
    x = expression("Equivalent uncontrolled points"~hat(x)[u]),
    y = "Risk"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )
```


